# Analysis of Baldock and Souza networks
In this vignette we'll analyze an extract of the Doré et al. dataset. Namely networks from Baldock and some from Souza.

```{r}
#| echo = TRUE,
#| message = FALSE
library(colSBM)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)
library(here)
```

```{r}
# Load the data
all_dore_matrices <- readRDS(here("data", "dore-binary-matrices.Rds"))
baldock_matrices <- all_dore_matrices[grepl("Baldock", x = names(all_dore_matrices))]
souza_matrices <- all_dore_matrices[grepl("Souza", x = names(all_dore_matrices))]
```

We remarked that interesting results were obtained when clustering networks with close density.

```{r}
bipartite_graph_density <- function(A) {
    # Calculate the density of a bipartite graph
    # A: adjacency matrix of the bipartite graph
    # Returns the density of the bipartite graph

    n <- nrow(A)
    m <- ncol(A)
    d <- sum(A) / (n * m)
    return(d)
}

baldock_densities <- sapply(baldock_matrices, bipartite_graph_density)
souza_densities <- sapply(souza_matrices, bipartite_graph_density)
densities_df <- data.frame(
    name = c(names(baldock_densities), names(souza_densities)),
    density = c(baldock_densities, souza_densities),
    type = c(rep("Baldock", length(baldock_densities)), rep("Souza", length(souza_densities)))
)
```

The densities are as follows:
```{r}
#| echo = FALSE, fig.cap = "Density of Baldock and Souza networks"
ggplot(densities_df, aes(x = name, y = density, fill = type)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    labs(title = "Density of Baldock and Souza networks", x = "Network", y = "Density") +
    theme_minimal()
```

As they are close, we can analyze them together. 

### Souza networks
```{r}
#| cache=TRUE
set.seed(123)
if (!file.exists(here("data", "souza_fit.Rds"))) {
    souza_fit <- colSBM::estimate_colBiSBM(
        netlist = souza_matrices,
        net_id = names(souza_matrices),
        colsbm_model = "iid",
        global_opts = list(backend = "no_mc"),
        fit_opts = list(kmeans_nstart = 500L, max_vem_steps = 5000L)
    )
    saveRDS(souza_fit, here("data", "souza_fit.Rds"))
} else {
    souza_fit <- readRDS(here("data", "souza_fit.Rds"))
}
souza_fit$choose_joint_or_separated()
```

```{r}
plot(souza_fit) + plot(souza_fit$best_fit, type = "meso", values = TRUE)
```

### Baldock networks
```{r}
if (!file.exists(here("data", "baldock_fit.Rds"))) {
    set.seed(123)
    baldock_fit <- colSBM::estimate_colBiSBM(
        netlist = baldock_matrices,
        net_id = names(baldock_matrices),
        colsbm_model = "iid",
        global_opts = list(backend = "no_mc"),
        fit_opts = list(kmeans_nstart = 500L, max_vem_steps = 5000L)
    )
    saveRDS(baldock_fit, here("data", "baldock_fit.Rds"))
} else {
    baldock_fit <- readRDS(here("data", "baldock_fit.Rds"))
}
baldock_fit$choose_joint_or_separated()
```

```{r}
plot(baldock_fit) + plot(baldock_fit$best_fit, type = "meso", values = TRUE)
```

### Joint analysis of Baldock and Souza networks

```{r}
#| cache=TRUE
if (!file.exists(here("data", "baldock_and_souza_fit.Rds"))) {
    set.seed(123)
    baldock_and_souza_fit <- colSBM::estimate_colBiSBM(
        netlist = c(baldock_matrices, souza_matrices),
        net_id = c(names(baldock_matrices), names(souza_matrices)),
        colsbm_model = "iid",
        global_opts = list(backend = "no_mc"),
        fit_opts = list(kmeans_nstart = 500L, max_vem_steps = 5000L)
    )
    saveRDS(baldock_and_souza_fit, here("data", "baldock_and_souza_fit.Rds"))
} else {
    baldock_and_souza_fit <- readRDS(here("data", "baldock_and_souza_fit.Rds"))
}

baldock_and_souza_fit$choose_joint_or_separated()
```

```{r}
plot(baldock_and_souza_fit) + plot(baldock_and_souza_fit$best_fit, type = "meso", values = TRUE)
``` 

And to recapitulate the results, we can plot the best fit for each of the three analyses:
```{r}
#| echo = FALSE, fig.cap = "Best fit for Baldock, Souza and Baldock + Souza networks"
baldock_plot <- plot(baldock_fit$best_fit, type = "meso", values = TRUE) +
    labs(title = "Baldock networks")
souza_plot <- plot(souza_fit$best_fit, type = "meso", values = TRUE) +
    labs(title = "Souza networks")
baldock_and_souza_plot <- plot(baldock_and_souza_fit$best_fit, type = "meso", values = TRUE) +
    labs(title = "Baldock + Souza networks")
combined_plot <- baldock_plot + souza_plot + baldock_and_souza_plot +
    plot_layout(ncol = 3) +
    plot_annotation(title = "Best fit for Baldock, Souza and Baldock + Souza networks")
combined_plot
```

### Clustering Baldock

```{r}
if (!file.exists(here("data", "baldock_clustering.Rds"))) {
    baldock_clustering <- colSBM::clusterize_bipartite_networks(
        netlist = baldock_matrices,
        net_id = names(baldock_matrices),
        colsbm_model = "iid",
        global_opts = list(backend = "no_mc"),
        fit_opts = list(kmeans_nstart = 500L, max_vem_steps = 5000L)
    )
    saveRDS(baldock_clustering, here("data", "baldock_clustering.Rds"))
} else {
    baldock_clustering <- readRDS(here("data", "baldock_clustering.Rds"))
}
print(
    paste0(
        "BIC-L of Baldock clustering separating African network from English ones: ",
        sum(sapply(baldock_clustering$partition, function(x) x$BICL))
    )
)
print(paste0(
    "BIC-L of all Baldock networks: ",
    baldock_fit$best_fit$BICL
))
```

Thus separating the African network from the English ones is a better fit than clustering all Baldock networks together.

## Clustering Souza
```{r}
if (!file.exists(here("data", "souza_clustering.Rds"))) {
    souza_clustering <- colSBM::clusterize_bipartite_networks(
        netlist = souza_matrices,
        net_id = names(souza_matrices),
        colsbm_model = "iid",
        global_opts = list(backend = "no_mc"),
        fit_opts = list(kmeans_nstart = 500L, max_vem_steps = 5000L)
    )
    saveRDS(souza_clustering, here("data", "souza_clustering.Rds"))
} else {
    souza_clustering <- readRDS(here("data", "souza_clustering.Rds"))
}
print(
    paste0(
        "BIC-L of Souza clustering: ",
        sum(sapply(souza_clustering$partition, function(x) x$BICL))
    )
)
print(paste0(
    "BIC-L of all Souza networks: ",
    souza_fit$best_fit$BICL
))
```

This indicates that jointly analyzing the Souza networks or clustering is marganally different.

### Clustering Baldock and Souza
```{r}
options(future.global.maxsize = 10^9)
if (!file.exists(here("data", "baldock_and_souza_clustering.Rds"))) {
    baldock_and_souza_clustering <- colSBM::clusterize_bipartite_networks(
        netlist = c(baldock_matrices, souza_matrices),
        net_id = c(names(baldock_matrices), names(souza_matrices)),
        colsbm_model = "iid",
        global_opts = list(backend = "no_mc"),
        fit_opts = list(kmeans_nstart = 500L, max_vem_steps = 5000L)
    )
    saveRDS(baldock_and_souza_clustering, here("data", "baldock_and_souza_clustering.Rds"))
} else {
    baldock_and_souza_clustering <- readRDS(here("data", "baldock_and_souza_clustering.Rds"))
}
print(
    paste0(
        "BIC-L of Baldock + Souza clustering: ",
        sum(sapply(baldock_and_souza_clustering$partition, function(x) x$BICL))
    )
)
print(paste0(
    "BIC-L of all Baldock + Souza networks: ",
    baldock_and_souza_fit$best_fit$BICL
))
```

But we want to see if the BIC-L is better for the Baldock networks, let's extract their variational bound (as penalty is computed over all networks).

```{r}
extract_vbound_per_net <- function(fit) {
    setNames(sapply(seq_along(fit$A), function(m) {
        fit$vb_tau_pi(m) + fit$vb_tau_alpha(m) + fit$entropy_tau(m)
    }), fit$net_id)
}

baldock_and_souza_vbound <- unlist(lapply(baldock_and_souza_clustering$partition, function(x) extract_vbound_per_net(x)))
baldock_vbound <- unlist(sapply(baldock_clustering$partition, function(x) extract_vbound_per_net(x)))
souza_vbound <- unlist(lapply(souza_clustering$partition, function(x) extract_vbound_per_net(x)))

baldock_vs_baldock_and_souza_vbound <- baldock_vbound - baldock_and_souza_vbound[grepl("Baldock", names(baldock_and_souza_vbound))]
souza_vs_baldock_and_souza_vbound <- souza_vbound - baldock_and_souza_vbound[grepl("Souza", names(baldock_and_souza_vbound))]
```
The variational bound for Baldock and Souza together is $`r sum(baldock_and_souza_vbound[grepl("Baldock", names(baldock_and_souza_vbound))])`$ but for Baldock clustering only it is $`r 
sum(baldock_vbound)`$ this indicates that Baldock networks are better fitted when clustering them together but without adding Souza networks.

| Authors clustered | BIC-L | Variational bound |
|------------------|-------|------------------|
| Baldock          | `r sum(sapply(baldock_clustering$partition, function(x) x$BICL))` | `r sum(baldock_vbound)` |
| Souza            | `r sum(sapply(souza_clustering$partition, function(x) x$BICL))` | `r sum(souza_vbound)` |
| Baldock + Souza  | `r sum(sapply(baldock_and_souza_clustering$partition, function(x) x$BICL))` | `r sum(baldock_and_souza_vbound)` |

## Achieving best partitionning 

We've seen that directly clustering Baldock networks is better than clustering them with Souza networks. 
We will now try to achieve the best partitionning by initializing the clustering with the best fit of Baldock and of Souza networks.
But to see if the clustering is capable of cutting African network from Baldock English ones, we will use the best fit of Baldock and the best fit for Souza.

```{r}
# options(future.globals.maxSize = 10^9)
# baldock_and_souza_from_prefit <- clusterize_bipartite_networks(
#     netlist = c(baldock_matrices, souza_matrices),
#     net_id = c(names(baldock_matrices), names(souza_matrices)),
#     colsbm_model = "iid",
#     global_opts = list(backend = "no_mc"),
#     fit_opts = list(kmeans_nstart = 500L, max_vem_steps = 5000L),
#     fit_init = list(
#         baldock_fit,
#         souza_fit
#     )
# )
```

Separating African and starting from Baldock English union Souza

```{r}
baldock_eng_matrices <- baldock_matrices[-1]
baldock_eng_and_souza <- estimate_colBiSBM(
    netlist = c(baldock_eng_matrices, souza_matrices),
    net_id = c(names(baldock_eng_matrices), names(souza_matrices)),
    colsbm_model = "iid",
    global_opts = list(backend = "no_mc"),
    fit_opts = list(kmeans_nstart = 500L, max_vem_steps = 5000L)
)
baldock_afr <- estimate_colBiSBM(
    netlist = baldock_matrices[1],
    net_id = names(baldock_matrices)[1],
    colsbm_model = "iid",
    global_opts = list(backend = "no_mc"),
    fit_opts = list(kmeans_nstart = 500L, max_vem_steps = 5000L)
)
```